Using Random Walk Simulations to Calculate Ground State Energies in Quantum Physics

### 2020-04-19 Paper/main.tex
Rewrote Conclusions sections

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
% Provide a summary of the session. What was done, what measurements were taken,
% brief methods, what calculations, brief conclusion.  The Abstract should be
% approximately 250 words or fewer, italicized, in 10-point Times (or Times
% Roman.) Please leave two spaces between the Abstract and the heading of your
% first section.  It should briefly summarize the essence of the paper and address
% the following areas without using specific subsection titles. Objective: Briefly
% state the problem or issue addressed, in language accessible to a general
% scientific audience. Technology or Method: Briefly summarize the technological
% innovation or method used to address the problem. Results: Provide a brief
% summary of the results and findings. Conclusions: Give brief concluding remarks
% on your outcomes. Detailed discussion of these aspects should be provided in the
% main body of the paper.
\end{abstract}

@@

In this investigation, I explore an alternative Monte Carlo method using Random
Walk simulations, and demonstrate its efficacy with simple systems.

@@

\begin{figure}%[H]%%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/cartoon.pdf}
    \caption{The position of a walker at each step in a 20-step random walk
      simulation. The average position of the walker does not necessarily remain
      at the origin as one might naively expect. Instead it is normal for the
      walker to finish the walk at some distance from the origin, as shown here.}
    \label{fig:cartoon}
  \end{center}
\end{figure}

@@

This investigation applied Random walk simulations to calculate the ground state
energies of Quantum Systems by solving the Schr\"{o}dinger equation, as shown in
\todo{cite}.

@@

\begin{equation}
  \label{eq:squarewell}
    a(j) =
    \begin{cases}
      0,& \text{if } j < J\\
      \infty,& \text{if } j \geq J
    \end{cases}
    \nonumber
\end{equation}
where J is a well-defined boundary. So the walker cannot die inside the well but
wil immediately die when they step outside, which we expect will definitely
happen given sufficient time. Inside the well, the potential is simply that of a
free particle \todo{cite}:

@@

\begin{equation}
  E = \lambda R^2 \frac{\hbar^2}{mJ^2}
  \nonumber
\end{equation}
where only $\lambda$ is an unknown quantity. Thus, by calculating rate of death
$\lambda$, we can calculate the ground state energy of the infinite square
well. The known ground state energy of an infnite square well is
$\frac{\pi^2}{8}\frac{\hbar^2}{mJ^2}$ \todo{cite}, so we expect to find $\lambda
R^2$ close to $\frac{\pi}{8}$.

@@

\begin{figure*}%[ht!]
  \centering
  \subfloat[Histogram of probability of surviving to at least a particlar step]
  {\includegraphics[width=0.5\textwidth]{images/cum_exp_plot.pdf}
  \label{fig:cumhist}}
  \centering
  \subfloat[Histogram of probability of surviving to at least a particular step
  in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot.pdf}
  \label{fig:cumloghist}}
  \caption{Histogram showing the Probability of surviving to at least a
    particular step in an infinite square well with $J = 20$. The peak of 1 does
    not start at the origin but at step 100. This is because these steps were
    neglected in the plot as prior to this step the exponential shape is less
    clear, as explained by Figure \ref{fig:hist}. An exponential decay curve
    provides a good fit, with an $R^2$ value of 0.9999998, very close to
    1. Plotting on a logarithmic scale shows the straight line fit, the gradient
    of which is $\lambda$, giving $\lambda J^2 = 1.2334$, with a negligible
    error. The fluctuation in the higher step number regions is due to the large
    variance in the small number of walkers that die at these steps.}
  \label{fig:cumplots}
\end{figure*}

@@

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/exp_plot.pdf}
    \caption{The probability of dying on a particular step in an infnite square
      well with $J=20$. As can be seen, the peak is away from the origin. This
      is because, for a well with boundary $J$, the minimum number of steps a
      walker must take is $J$. However, it is still unlikely that the walker
      will die on this minimum number of steps, as then the walker must walk in
      the same direction every time. Thus, the most probable step for the walker
      to die on is greater than $J$. It is for this reason that the exponential
      decay does not begin at the origin in Figure \ref{fig:cumhist}, meaning
      some earlier steps must be neglected in order to achieve a better fit.}
    \label{fig:hist}
  \end{center}
\end{figure}

@@

\begin{figure*}%[ht!]
  \centering
  \subfloat[Probability of dying on a particlar step with a step cutoff]
  {\includegraphics[width=0.5\textwidth]{images/exp_plot_cutoff.pdf}
  \label{fig:expcutoff}}
  \centering
  \subfloat[Probability of surviving to at least a particular step in a
  logarithmic scale with a step cutoff]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot_cutoff.pdf}
  \label{fig:cum_line_cutoff_plot}}
  \caption{Histograms showing the results of the simulation when a maximm number
  of steps is introduced. If a walker reaches this number of steps, they
  automatically die, regardless of position. This results in a
  disproportionately high number of walkers dying at this maximum number, 2000
  in this case, as shown by Figure \ref{fig:expcutoff}. However, this results in
  no fluctuation at higher number of steps, as shown in Figure
  \ref{fig:cum_line_cutoff_plot}, though some systematic error is introduced
  through this method.}
  \label{fig:cutoffplots}
\end{figure*}

@@

@@

Running the simulation 10 times and taking the average $\lambda J^2$ with error
being standard deviation, we find:
\begin{equation}
  \lambda J^2 = 1.2339 \pm 0.0008
  \nonumber
\end{equation}
We quote the precision to the decimal place at which there is variation from the
theoretical result of $1.2337$. We notice that the error in the calculated
answer is small (since there is not much variation between repeated simulations)
and the theoretical result is within the range of the calculated answer.

\todo{circular well}

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/multiplot1.pdf}
    \caption{Histogram plotted on a logarithmic scale showing the Probability of
      surviving to at least a particualr step in different infinite square wells
      with different boundaries $J$. The simulations were run with $10000$
      walkers The $\lambda J^2$ values for each are shown. The values get closer
      to the theoretical value of $\frac{\pi^2}{8} = 1.2337$ as we increase the
      size of the well. This is likely due to more walkers surviving to a higher
      number of steps, as the well size is increased.}
    \label{fig:multi_line_plot}
  \end{center}
\end{figure}

@@

In this investigation we demonstrated a method of calculating ground state
energies of a quantum system using Random walk computational simulations. The
investigation focussed on an Infinite Square Well, but is easily extended to a
Symmetric potential. The simualation was written in C++ to maximise performance
and is available at \todo{cite}.

The investigation explored the effects of varying the size of the well in
determining the ground state energy of a quantum system, and showed that this
method is more effective for larger wells, i.e. wells with larger
boundaries. This is because for wells with larger boundaries, the walker can
take a larger number of steps on average before dying, and thus can populate
larger step bins. Thus, this method produces data that is close to the
theoretical result for larger infinite square wells.

The general theoretical result for the ground state energy of an infinite square
well is $\frac{\pi^2}{8}\frac{\hbar^2}{mJ^2}$ where $J$ is the boundary of the
well, and this method allows for the calculation of $\lambda R^2
\frac{\hbar^2}{mJ^2}$, where $R$ also corresponds to the boundary of the
well. Thus, we expect our calculated result to be $\lambda R^2 = \lambda J^2 =
\frac{\pi^2}{8}$.

@@

\appendices
\section{Random walk Schr\"{o}dinger Equation Derivation}
\label{appendix:derivation}
Consider a one-dimensional lattice upon which a walker is undergoing a random
walker. Let us denote the probability that the walker is at position $j$ after
$n$ steps as $p(j,n)$ and the probabilty that the walker will die at position
$j$ as $a(j)$. We will denote the position-step number coordinates as $[j,
n]$. Thus, for a walker to have coordinates $[j, n+1]$, in the previous step,
they must have had coordinates $[j \pm 1, n]$. So we can say:

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
% Provide a summary of the session. What was done, what measurements were taken,
% brief methods, what calculations, brief conclusion.  The Abstract should be
% approximately 250 words or fewer, italicized, in 10-point Times (or Times
% Roman.) Please leave two spaces between the Abstract and the heading of your
% first section.  It should briefly summarize the essence of the paper and address
% the following areas without using specific subsection titles. Objective: Briefly
% state the problem or issue addressed, in language accessible to a general
% scientific audience. Technology or Method: Briefly summarize the technological
% innovation or method used to address the problem. Results: Provide a brief
% summary of the results and findings. Conclusions: Give brief concluding remarks
% on your outcomes. Detailed discussion of these aspects should be provided in the
% main body of the paper.
  In this paper, we present a novel way of calculating ground state energies for
  a quantum simulation using a Random Walk Monte Carlo simulation.
\end{abstract}

@@

In this investigation, we explore an alternative Monte Carlo method using Random
Walk simulations, and demonstrate its efficacy with simple systems.

@@

\begin{figure}[H]%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/cartoon.pdf}
    \caption{The position of a walker at each step in a 20-step random walk
      simulation. The average position of the walker does not necessarily remain
      at the origin as one might naively expect. Instead it is normal for the
      walker to finish the walk at some distance from the origin, as shown here.}
    \label{fig:cartoon}
  \end{center}
\end{figure}

@@

This investigation applied Random walk simulations to calculate the ground state
energies of Quantum Systems by solving the Schr\"{o}dinger equation, as shown in
\cite{Ceperley1983}.

@@

\begin{equation}
  \label{eq:squarewell}
    a(j) =
    \begin{cases}
      0,& \text{if } j < J\\
      \infty,& \text{if } j \geq J
    \end{cases}
    \nonumber
\end{equation}
where J is a well-defined boundary. So the walker cannot die inside the well but
wil immediately die when they step outside, which we expect will definitely
happen given sufficient time. Inside the well, the potential is simply that of a
free particle \cite{Bransden2003}:

@@

\begin{equation}
  E = \lambda R^2 \frac{\hbar^2}{mJ^2}
  \nonumber
\end{equation}
where only $\lambda$ is an unknown quantity. Thus, by calculating rate of death
$\lambda$, we can calculate the ground state energy of the infinite square
well. The known ground state energy of an infnite square well is
$\frac{\pi^2}{8}\frac{\hbar^2}{mJ^2}$ \cite{Davies1998}, so we expect to find $\lambda
R^2$ close to $\frac{\pi}{8}$.

@@

\begin{figure*}[ht!]
  \centering
  \subfloat[Histogram of probability of surviving to at least a particlar step]
  {\includegraphics[width=0.5\textwidth]{images/cum_exp_plot.pdf}
  \label{fig:cumhist}}
  \centering
  \subfloat[Histogram of probability of surviving to at least a particular step
  in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot.pdf}
  \label{fig:cumloghist}}
  \caption{Histogram showing the Probability of surviving to at least a
    particular step in an infinite square well with $J = 20$. The peak of 1 does
    not start at the origin but at step 100. This is because these steps were
    neglected in the plot as prior to this step the exponential shape is less
    clear, as explained by Figure \ref{fig:hist}. An exponential decay curve
    provides a good fit, with an $R^2$ value of 0.9999998, very close to
    1. Plotting on a logarithmic scale shows the straight line fit, the gradient
    of which is $\lambda$, giving $\lambda J^2 = 1.2334$, with a negligible
    error. The fluctuation in the higher step number regions is due to the large
    variance in the small number of walkers that die at these steps.}
  \label{fig:cumplots}
\end{figure*}

@@

\begin{figure}[H]%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/exp_plot.pdf}
    \caption{The probability of dying on a particular step in an infnite square
      well with $J=20$. As can be seen, the peak is away from the origin. This
      is because, for a well with boundary $J$, the minimum number of steps a
      walker must take is $J$. However, it is still unlikely that the walker
      will die on this minimum number of steps, as then the walker must walk in
      the same direction every time. Thus, the most probable step for the walker
      to die on is greater than $J$. It is for this reason that the exponential
      decay does not begin at the origin in Figure \ref{fig:cumhist}, meaning
      some earlier steps must be neglected in order to achieve a better fit.}
    \label{fig:hist}
  \end{center}
\end{figure}

@@

\begin{figure*}[ht!]
  \centering
  \subfloat[Probability of dying on a particlar step with a step cutoff]
  {\includegraphics[width=0.5\textwidth]{images/exp_plot_cutoff.pdf}
  \label{fig:expcutoff}}
  \centering
  \subfloat[Probability of surviving to at least a particular step in a
  logarithmic scale with a step cutoff]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot_cutoff.pdf}
  \label{fig:cum_line_cutoff_plot}}
  \caption{Histograms showing the results of the simulation when a maximm number
  of steps is introduced. If a walker reaches this number of steps, they
  automatically die, regardless of position. This results in a
  disproportionately high number of walkers dying at this maximum number, 2000
  in this case, as shown by Figure \ref{fig:expcutoff}. However, this results in
  no fluctuation at higher number of steps, as shown in Figure
  \ref{fig:cum_line_cutoff_plot}, though some systematic error is introduced
  through this method.}
  \label{fig:cutoffplots}
\end{figure*}

@@

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/multiplot1.pdf}
    \caption{Histogram plotted on a logarithmic scale showing the Probability of
      surviving to at least a particular step in different infinite square wells
      with different boundaries $J$. The simulations were run with $10000$
      walkers The $\lambda J^2$ values for each are shown. The values get closer
      to the theoretical value of $\frac{\pi^2}{8} = 1.2337$ as we increase the
      size of the well. This is likely due to more walkers surviving to a higher
      number of steps, as the well size is increased.}
    \label{fig:multi_line_plot}
  \end{center}
\end{figure}

@@

Running the simulation 10 times with 1000000 walkers in an infinite square well
with $J=20$ and taking the average $\lambda J^2$ with error being standard
deviation, we find:
\begin{equation}
  \lambda J^2 = 1.2339 \pm 0.0008
  \nonumber
\end{equation}
We quote the precision to the decimal place at which there is variation from the
theoretical result of $1.2337$. We notice that the error in the calculated
answer is small (since there is not much variation between repeated simulations)
and the theoretical result is within the range of the calculated answer.

@@

In this investigation we demonstrated a method of calculating ground state
energies of a quantum system using Random walk computational simulations. The
investigation focussed on a one dimensional infinite square well, but can easily
be extended to more complex systems such as a circular well in two dimensions.

The investigation explored the effects of varying the size of the well in
determining the ground state energy of a quantum system, and showed that this
method produces more accurate results for ground state energies of larger wells.
This is because for wells with larger boundaries, the walker can take a larger
number of steps on average before dying, and thus can populate larger step
bins. Thus, this method produces data that is close to the theoretical result
for larger infinite square wells.

The general theoretical result for the ground state energy of an infinite square
well is $\frac{\pi^2}{8}\frac{\hbar^2}{mJ^2}$ where $J$ is the boundary of the
well, and this method allows for the calculation of $\lambda R^2
\frac{\hbar^2}{mJ^2}$, where $R$ also corresponds to the boundary of the
well. Thus, we expect our calculated result to be $\lambda J^2 =
\frac{\pi^2}{8}$.

@@

\pagebreak
\appendices
\section{Random walk Schr\"{o}dinger Equation Derivation}
\label{appendix:derivation}
Consider a one-dimensional lattice upon which a walker is undergoing a random
walker. Let us denote the probability that the walker is at position $j$ after
$n$ steps as $p(j,n)$ and the probabilty that the walker will die at position
$j$ as $a(j)$. We will denote the position-step number coordinates as $[j,
n]$. Thus, for a walker to have coordinates $[j, n+1]$, in the previous step,
they must have had coordinates $[j \pm 1, n]$. So we can say:
